A framework for pipeline benchmarking and its application to single-cell RNAseq clustering

### 2020-01-30 bmc_article.tex
Updates from Overleaf

While a number of good comparison and benchmark studies have already been performed on various steps related to scRNAseq analysis and can inform analytical decisions (\citealp{duoClustering2018, SonesonDE2018, SunDimRed2019}), they need constant updating and often leave open many details of an analysis. Another missing aspect of current benchmarking studies is their limitation to capture all aspects of scRNAseq processing workflow. Although previous benchmarks already brought valuable recommendations for data processing, some only focused on one aspect of data processing (e.g., \citealp{SunDimRed2019}), did not evaluate the effect of the choice of a tool on downstream analysis (e.g., \citealp{TsuyuzakiPCA2020}) or did not tackle all aspects of data processing, such as doublet identification or cell filtering (e.g., \citealp{viethSystematic2019}). A thorough evaluation of the tools covering all major processing steps is however urgently needed as previous benchmarking studies highlighted that a combination of tools can have a drastic impact on downstream analysis, such as differential expression analysis and cell-type deconvolution (\citealp{viethSystematic2019, CobosDeconvolution2020}). It is then likely that processing algorithms also interact in a positive or negative fashion and it is then critical to evaluate not only the single effect of a method but also its interaction with all parts of a workflow.

Here, we develop a flexible R framework for pipeline comparison and evaluate the various steps of analysis leading from an initial count matrix to a cluster assignment, which are critical in a wide range of applications. We harnessed real datasets of known cell composition (Table \ref{tab:table1}) and develop new evaluation metrics to investigate in a multilevel fashion the impact of various parameters and variations around a core scRNAseq pipeline. Although we use some datasets based on other protocols, our focus is especially on droplet-based datasets that do not include exogenous control RNA (i.e. spike-ins); see Table \ref{tab:table1} and Figure \ref{fig:figure1} for a description of the datasets. In addition to previously-used benchmark datasets with true cell labels \cite{duoClustering2018,tianMixology2018}, we simulated two datasets with a hierarchical subpopulation structure based on real 10x human and mouse data using \textit{muscat} \cite{CrowellMuscat2019}). 
Since graph-based clustering \cite{satijaSeurat2015} was previously shown to consistently perform well across several datasets \cite{duoClustering2018,tianMixology2018}, we used the \textit{Seurat} pipeline as the starting point to perform an integrated investigation of: 1) doublet identification, 2) cell filtering, 3) normalization, 4) feature selection, 5) dimension reduction, 6) clustering. We compare not only competing approaches, but also more fine-grained parameters and variations on common methods. Importantly, the success of methods at a certain analytical step might be dependent on choices at other steps. Therefore, instead of evaluating each step in isolation, we developed a general framework for evaluating nested variations on a pipeline, and suggest a multilevel panel of metrics. Finally, we evaluate several recent methods and provide concrete recommendations.

@@

The \textit{pipeComp} package defines a pipeline as, minimally, a list of functions executed consecutively on the output of the previous one (Figure \ref{fig:figure2}A). In addition, optional benchmark functions can be set for each step to provide standardized, multi-layered evaluation metrics. Given such a \texttt{PipelineDefinition} object, a set of alternative parameters (which might include different subroutines) and benchmark datasets, the \texttt{runPipeline} function then proceeds through all combinations of arguments, avoiding recomputing the same step twice and compiling evaluations (including running time) on the fly. The \textit{CellBench} package was recently proposed to address a similar need\citep{su_cellbench}, offering an elegant piping syntax to combine alternative methods at successive steps. In addition to offering the same capabilities, a key additional feature of \textit{pipeComp} is that, since the evaluation metrics are not treated as a step but parallel to it, they can be performed at various points in the pipeline without needing to store the potentially large intermediate data, and thereby enabling combinatorial benchmarks. Variations in a given parameter can be evaluated using all metrics from this point downward in the pipeline. This is especially important because end-point metrics, such as the adjusted Rand index (ARI, \citep{HubertARI1985}) for clustering, are not perfect. For example, by far the most important determinant of the ARI is the number of detected clusters: the farther it is from the actual number of subpopulations, the worse the ARI. In this context, one strategy has been to cluster across various resolutions and only consider the results that have the right number of clusters \citep{duoClustering2018}. While this has the virtue of making the results comparable, in practice the number of subpopulations is typically unknown, and tools that operate well in this optimal context might not necessarily be best overall. Clustering results are also very sensitive, and might not always capture improvements in earlier steps of the pipeline.

@@

Mitochondrial reads have been associated with cell degradation and there is evidence that ribosomal genes can influence the clustering output, hiding other biological structure in the analysis \citep{freytagComparison2018}. We therefore investigated whether excluding one category of features or the other, or using only protein-coding genes, had an impact on the ability to distinguish subpopulations (Supplementary Figure 9). Removal of ribosomal genes robustly reduced the quality of the clustering, suggesting that they represent real biological differences between subpopulations. Removing mitochondrial genes and restricting to protein-coding genes had a very mild impact.

@@

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{{main_figures/normalization_files/figure-html/figure3-1.png}}
    \caption{\textbf{Evaluation of normalization procedures.} `Seurat.feat\_mt\_regress' indicates the regressing out of the number of features and proportion of mitochondrial reads during scaling, represents, and `Seurat.feat\_regress` of the number of features only.  Residual correlation of the first principal component with library size (left) and detection rate (right), after accounting for biological differences between subpopulations. \textbf{B:} Average silhouette width per true subpopulation, where higher silhouette width means a higher separability. \textbf{C:} Clustering accuracy, measured by the ARI at the true number of cluster (left) and by the average mutual information (MI) of the cluster assignment with true subpopulations.}
    \label{fig:figure5}
\end{figure}

@@

We next investigated the methods of dimensionality reduction. Since the various PCA approaches and implementations were recently benchmarked in a similar context \citep{TsuyuzakiPCA2020}, we focused on widely used approaches which had not yet been compared: \textit{Seurat}'s PCA, \textit{scran}'s \texttt{denoise PCA}, and GLM-PCA \citep{townesGlmpca2019}, and combined them (when relevant) with \textit{sctransform} normalization. Given that \textit{Seurat}'s default PCA weights the cell embeddings by the variance of each component, we also evaluated the impact of this weighting with each method.

The impact of the choice of dimensionality reduction method was far greater than that of normalization or feature selection (e.g. Supplementary Figure 15). The first components identified by GLM-PCA tended to have a higher proportion of their variance explained by real subpopulations and increase the average silhouette width of already well-defined subpopulations. Seurat's PCA procedure however proved superior on all metrics (Figure \ref{fig:figure7}). 
Overall, weighting the principal components by their variance, in the same fashion as Seurat's method, had a strong positive impact on silhouette widths and ARI scores.

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{{main_figures/dr_files/figure-html/dimred-1.png}}
    \caption{\textbf{Evaluation of common dimensionality reduction methods. A:} Average silhouette width per subpopulation resulting from combinations of normalization and dimension reductions. \textbf{B:} Clustering accuracy , mean proportion of the variance in the first 5 components explained by real subpopulations (center), and median ARI of the resulting clustering (right).}
    \label{fig:figure7}
\end{figure}

@@

A common step following dimension reduction is the selection of an appropriate number of dimensions to use for downstream analysis. Since Euclidean distance decreases as the number of non-discriminating dimensions increases, there is usually a trade-off between selecting enough dimensions to keep most information and excluding smaller dimensions that may represent technical noise or other unwanted sources of variation. Overall, increasing the number of dimensions robustly led to a decrease in the number of clusters (Supplementary Figures 15). This tended to be translated to the accuracy of the clustering (Supplementary Figure 16), although in both cases (number of clusters and ARI) \textit{Seurat}'s resolution parameter had a much stronger impact.

@@

Since ARI is dominated by differences in the number of clusters, and no single metric is perfect, we diversified them (Figure \ref{fig:figure9}). Mutual information (MI) has the virtue of not decreasing when a true subpopulation is split into two clusters, which is arguably less problematic (and might well reflect unknown biological subgroups), but as a consequence it can be biased towards methods producing higher resolution clustering. Similarly, precision per true subpopulation is considerably more robust to differences in the number of clusters. We also tracked the mean F1 score and the ARI at the true number of clusters.

The MI score and minimum precision, which are largely independent of the estimated number of clusters, were overall higher for the \textit{walktrap} method (Figure \ref{fig:figure9}), while the mean F1 score favored both \textitt{scran} methods (\textit{walktrap} and \textit{fast greedy}) over \textit{Seurat}. Finally, the ARI score at the true number of clusters, when available, showed similar performances. However, while \textit{Seurat}'s resolution parameter had a large impact on the number of clusters identified, and therefore could be used to find a parameter combination producing the right number of clusters, the number of clusters found by \textit{scran}-based clustering was considerably less influenced by the available parameters (number of nearest neighbors or steps in the random walk - see Supplementary Figure 17). The occasional lack of scran-based clustering with the right number of clusters, along with a higher mutual information, suggest that \textit{scran} sometimes simply divides a real subpopulation into two clusters (possibly tracking some unknown biological differences) rather than committing misclassification errors. Overall, the \textit{walktrap} method appeared superior to the \textit{fast greedy} algorithm and was generally less prone to misclassification than \textit{Seurat} clustering, although the latter offered more control over the resolution. Of note, very difficult subpopulations, both from the Zhengmix8eq and simMix1 datasets, remained very inaccurately classified by all methods and in regards to all metrics.

@@

The basic pipeline presented here can be extended with additional analysis steps while keeping the same evaluation metrics. To demonstrate this, we evaluated various imputation or denoising techniques based on their impact on classification. Since preliminary analysis showed that all methods performed equally well or better on normalized data, we applied them after filtering and normalization, but before scaling and reduction. Although some of the methods (e.g. \textit{DRImpute\_process} and \textit{alra\_norm}) did improve the separability of the more elusive subpopulations, no method had a systematically positive impact on the average silhouette width across all subpopulations (Figure \ref{fig:figure10}A). When restricting ourselves to clustering analyses that yielded the `right' number of clusters, all tested methods improved classification  compared to a scenario with no imputation step (`none' label, Figure \ref{fig:figure10}B). However, the situation was not so straightforward with alternative metrics, where some methods (e.g. \textit{enhance}) consistently underperformed. 10X datasets, which are typically characterized by a lower per-cell coverage and feature detection rate, benefited more from imputation, and in this context, \textit{DRImpute} and \textit{dca} tended to show the best performance. On the contrary, imputing on normalized counts originating from Smart-seq technology was instead rather deleterious to the clustering accuracy. Our recommendation follows the previous benchmarking of imputation strategies performed by \citep{ZhangImput2018}. Although this previous study focused on trajectory inference and preservation of data structure as main metrics, \textit{DrImpute} also revealed to be among the top performers and was recommended for clustering analysis (\textit{alra} was however not included in this previous evaluation).

@@

\begin{enumerate}
   \item Filtering
   \begin{itemize}
     \item Doublet detection and removal is advised, and can be performed at little computing cost with software such as \textit{scDblFinder} or \textit{scds}.
     \item Distribution-based cell filtering fails to capture doublets, and should use relatively lenient cutoffs (e.g. 5 MADs, or 3 MADs in at least 2 distributions) to exclude poor-quality cells while avoiding bias.
     \item Features filtering based on feature type did not appear beneficial.
   \end{itemize}
   \item Normalization and scaling
   \begin{itemize}
     \item Most normalization methods tested yielded a fair performance, especially when combined with scaling, which tended to have a positive impact on clustering.
     \item \textit{sctransform} offered the best overall performance in terms of the separability of the subpopulations, as well as removing the effect of library size and detection rate.
     \item The common practice of regressing out cell covariates, such as the detection rate or proportion of mitochondrial reads nearly always had a negative impact, leading to increased correlation with covariates and decreased clustering accuracy. We therefore advise against this practice.
   \end{itemize}
   \item Feature selection
   \begin{itemize}
     \item Deviance \citep{townesGlmpca2019} offered the best ranking of genes for feature selection.
     \item Increasing the number of features included tended to lead to better classifications, plateauing from 4000 features.
   \end{itemize}
   \item Denoising/imputation
   \begin{itemize}
       \item Denoising appeared beneficial to the identification of subpopulations in 10x datasets, but not in Smart-seq datasets.
       \item We found especially \textit{alra} (with prior normalization), \textit{DrImpute} (with prior processing) and \textit{dca} to offer the best performances.
   \end{itemize}
   \item PCA
   \begin{itemize}
     \item Similarly to previous reports \citep{SunDimRed2019}, we recommend the irlba-based PCA using weighting of the components, as implemented in \textit{Seurat}. 
     \item Instead of the common elbow or jackstraw methods for deciding on the components to include, we recommend the \textit{global maximum likelihood based on translated Poisson Mixture Model} method (e.g. implemented in \texttt{intrinsicDimension::maxLikGlobalDimEst()}).
   \end{itemize}
   \item Clustering
   \begin{itemize}
    \item We found \textit{scran}-based \textit{walktrap} clustering to show good performances.
    \item In cases where prior knowledge can guide the choice of a resolution, \textit{Seurat} can be useful in affording manual control of it while, in the absence of such knowledge, \textit{scran}-based \textit{walktrap} clustering provided reasonable estimates.
   \end{itemize}
\end{enumerate}

@@

In this study, we evaluated tools commonly used for the processing of scRNAseq with a focus on droplet-based datasets, namely from the 10x technology. Although this platform has been used in almost half of the scRNAseq studies in 2019 \citep{SvenssonDB2019}, other popular technologies such as Drop-seq, InDrops or Smart-seq2/3 were not represented in the present benchmarking. Differences between such protocols have a very large impact on clustering \citep{MereuCellAtlas2019}. Although most top-ranking methods in our comparison performed well on both Smart-seq and 10x datasets that we tested, future benchmarking efforts should strive to include less represented technologies. In addition, we did not compare any of the alignment and/or quantification methods used to obtain the count matrix \citep{MereuCellAtlas2019}. Some steps, such as the implementation of the PCA step, were also not explored in detail here as they have already been the object of recent and thorough study elsewhere \citep{SunDimRed2019}. We also considered only methods relying on Euclidean distance, while correlation was recently reported to be superior\citep{kim_impact_2019} and would require further investigation.

Here, we chose to concentrate on what could be learned from datasets with known cell labels (as opposed to labels inferred from the data, as in \citep{MereuCellAtlas2019}). In contrast to \citep{tianMixology2018}, who relied on RNA mixtures of known proportions, we chose to rely chiefly on real cells and their representative form of variability. Given the limited availability of such datasets however, several aspects of single-cell analysis could not be compared, such as batch effect correction or multi-dataset integration. For these aspects of scRNAseq processing that are critical in some experimental designs, we refer the reader to previous evaluations \citealp{TranBatch2020}. In addition, a focus on the identification of the subpopulations might fail to reveal methods that are instead best performing for tasks other than clustering, such as differential expression analysis or trajectory inference. Several informative benchmarks have already been performed on some of these topics \citealp{DalMolinDE2017, JaakkolaDE2017, SaelensTraject2019, SonesonDE2018}. Yet, such evaluations could benefit from considering methods not in isolation, but as parts of a connected workflow, as we have done here. We believe that the \textit{pipeComp} framework is modular and flexible enough to integrate new steps in the pipeline, as shown by an example with imputation/denoising methods.

@@

\textit{pipeComp} is a flexible R framework for evaluating methods and parameter combinations in a complex pipeline, computing on-the-fly multilevel evaluation metrics. Applying this framework to scRNAseq clustering enabled us to make concrete recommendations on the steps of filtering, normalization, feature selection, denoising, dimensionality reduction and clustering. We demonstrate how a diversity of multilevel metrics can be more robust, more sensitive, and more nuanced than simply evaluating the final clustering performance. In addition, we provide a new and efficient Bioconductor package for doublet detection, \texttt{scDblFinder}. We hope that the \textit{pipeComp} framework can be applied to extend the current field of benchmarking, as well as to apply it to other range of methods in scRNAseq analysis.

@@

\subsubsection*{Denoising/imputation}
Most imputation/denoising methods were run with the default parameters with the following exceptions; \textit{DrImpute} documentation advises to process the data prior to imputation by removing lowly expressed genes and cells expressing less than 2 genes. As it is not clear if this step is a hard requirement for the method to perform well, \textit{DrImpute} was run with and without prior processing (\textit{DrImpute\_process} and \textit{DrImpute\_noprocess} labels, respectively). \textit{dca} method does not accept non-integer counts, but two of the datasets used contained such table of counts due to multimapping. For these datasets, we rounded up the counts prior to imputation. \textit{alra} is designed for normalized data but as we are evaluating normalization downstream to imputation, we used the method on both non-normalized (\textit{alra} label) and normalized counts (\textit{alra\_norm} label). \textit{scImpute} requires an estimation of the expected number of clusters with the input data. As the estimation of the true number of cluster may not be known by the user, we evaluated the tool using the true number of clusters (\textit{scImpute} label) and using an over/under-estimation of this number (\textit{scImpute\_plus5} and \textit{scImpute\_min5} labels, respectively). \textit{ENHANCE} uses a k-nearest neighbor aggregation method and automatically estimates the number of neighbors to merge prior to the imputation. With the smallest datasets, this parameters was estimated to be 1, which lead to an early stop of the function. In such cases, we manually set this parameter to 2 for the method to work.

>>>>

While a number of good comparison and benchmark studies have already been performed on various steps related to scRNAseq analysis and can inform analytical decisions (\citealp{CobosDeconvolution2020, colePerformance2018, DalMolinDE2017, duoClustering2018, freytagComparison2018, HeiserDimred2019, JaakkolaDE2017, SonesonDE2018, SunDimRed2019, tianMixology2018, TranBatch2020, TsuyuzakiPCA2020, viethSystematic2019, WangDE2019, YipHVGs2018, ZhangImput2018}), they need constant updating and often leave open many details of an analysis. Another missing aspect of current benchmarking studies is their limitation to capture all aspects of scRNAseq processing workflow. Although previous benchmarks already brought valuable recommendations for data processing, some only focused on one aspect of data processing (e.g., \citealp{SunDimRed2019}), did not evaluate the effect of the choice of a tool on downstream analysis (e.g., \citealp{TsuyuzakiPCA2020}) or did not tackle all aspects of data processing, such as doublet identification or cell filtering (e.g., \citealp{viethSystematic2019}). A thorough evaluation of the tools covering all major processing steps is however urgently needed as previous benchmarking studies highlighted that a combination of tools can have a drastic impact on downstream analysis, such as differential expression analysis and cell-type deconvolution (\citealp{viethSystematic2019, CobosDeconvolution2020}). It is then likely that processing algorithms also interact in a positive or negative fashion and it is then critical to evaluate not only the single effect of a method but also its interaction with all parts of a workflow.

Here, we develop a flexible R framework for pipeline comparison and evaluate the various steps of analysis leading from an initial count matrix to a cluster assignment, which are critical in a wide range of applications. We harnessed real datasets of known cell composition (Table \ref{tab:table1}) and developed new evaluation metrics to investigate in a multilevel fashion the impact of various parameters and variations around a core scRNAseq pipeline. Although we use some datasets based on other protocols, our focus is especially on droplet-based datasets that do not include exogenous control RNA (i.e. spike-ins); see Table \ref{tab:table1} and Figure \ref{fig:figure1} for a description of the datasets. In addition to previously-used benchmark datasets with true cell labels \cite{duoClustering2018,tianMixology2018}, we simulated two datasets with a hierarchical subpopulation structure based on real 10x human and mouse data using \textit{muscat} \cite{CrowellMuscat2019}). 
Since graph-based clustering \cite{satijaSeurat2015} was previously shown to consistently perform well across several datasets \cite{duoClustering2018,tianMixology2018}, we used the \textit{Seurat} pipeline as the starting point to perform an integrated investigation of: 1) doublet identification, 2) cell filtering, 3) normalization, 4) feature selection, 5) dimension reduction, 6) clustering. We compare not only competing approaches, but also more fine-grained parameters and variations on common methods. Importantly, the success of methods at a certain analytical step might be dependent on choices at other steps. Therefore, instead of evaluating each step in isolation, we developed a general framework for evaluating nested variations on a pipeline, and suggest a multilevel panel of metrics. Finally, we evaluate several recent methods and provide concrete recommendations.

@@

The \textit{pipeComp} package defines a pipeline as, minimally, a list of functions executed consecutively on the output of the previous one (Figure \ref{fig:figure2}A). In addition, optional benchmark functions can be set for each step to provide standardized, multi-layered evaluation metrics. Given such a \texttt{PipelineDefinition} object, a set of alternative parameters (which might include different subroutines) and benchmark datasets, the \texttt{runPipeline} function then proceeds through all combinations of arguments, avoiding recomputing the same step twice and compiling evaluations (including running time) on the fly. The \textit{CellBench} package was recently proposed to address a similar need\citep{su_cellbench}, offering an elegant piping syntax to combine alternative methods at successive steps. In addition to offering the same capabilities, a key additional feature of \textit{pipeComp} is that, since the evaluation metrics are not treated as a step but parallel to it, they can be performed at various points in the pipeline without needing to store the potentially large intermediate data, and thereby enabling combinatorial benchmarks. Variations in a given parameter can be evaluated using all metrics from this point downward in the pipeline. This is especially important because end-point metrics, such as the adjusted Rand index (ARI, \citep{HubertARI1985}) for clustering, are not perfect. For example, although the meaning of an ARI score is independent of the number of true subpopulations \citep{steinley_properties_2004}, the number of clusters called is by far the most important determinant of the score: the farther it is from the actual number of subpopulations, the worse the ARI (Supplementary Figure 17). In this context, one strategy has been to cluster across various resolutions and only consider the results that have the right number of clusters \citep{duoClustering2018}. While this has the virtue of making the results comparable, in practice the number of subpopulations is typically unknown, and tools that operate well in this optimal context might not necessarily be best overall. Clustering results are also very sensitive, and might not always capture improvements in earlier steps of the pipeline.

@@

Mitochondrial reads have been associated with cell degradation, and there is evidence that ribosomal genes can influence the clustering output, hiding other biological structure in the analysis \citep{freytagComparison2018}. We therefore investigated whether excluding one category of features or the other, or using only protein-coding genes, had an impact on the ability to distinguish subpopulations (Supplementary Figure 9). Removal of ribosomal genes robustly reduced the quality of the clustering, suggesting that they represent real biological differences between subpopulations. Removing mitochondrial genes and restricting to protein-coding genes had a very mild impact.

@@

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{{main_figures/normalization_files/figure-html/figure3-1.png}}
    \caption{\textbf{Evaluation of normalization procedures.} `Seurat.feat\_mt\_regress' indicates the regressing out of the number of features and proportion of mitochondrial reads during scaling, represents, and `Seurat.feat\_regress` of the number of features only.  Residual correlation of the first principal component with library size (left) and detection rate (right), after accounting for biological differences between subpopulations. \textbf{B:} Average silhouette width per true subpopulation, where higher silhouette width means a higher separability. \textbf{C:} Clustering accuracy (\textit{Seurat} clustering), measured by the ARI at the true number of cluster (left) and by the average mutual information (MI) of the cluster assignment with true subpopulations.}
    \label{fig:figure5}
\end{figure}

@@

We next investigated the methods of dimensionality reduction. Since the various PCA approaches and implementations were recently benchmarked in a similar context \citep{TsuyuzakiPCA2020}, we focused on widely used approaches which had not yet been compared: \textit{Seurat}'s PCA, \textit{scran}'s \texttt{denoisePCA}, and GLM-PCA \citep{townesGlmpca2019}, and combined them (when relevant) with \textit{sctransform} normalization. Given that \textit{Seurat}'s default PCA weights the cell embeddings by the variance of each component, we also evaluated the impact of this weighting with each method.

The impact of the choice of dimensionality reduction method was far greater than that of normalization or feature selection (e.g. Supplementary Figure 15). GLM-PCA tended to increase the average silhouette width of already well-defined subpopulations, but \textit{Seurat}'s PCA procedure however proved superior on all metrics (Figure \ref{fig:figure7}). 
Overall, weighting the principal components by their variance (as \textit{Seurat} does) had a positive impact on silhouette widths and ARI scores.

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{{main_figures/dr_files/figure-html/dimred-1.png}}
    \caption{\textbf{Evaluation of common dimensionality reduction methods. A:} Average silhouette width per subpopulation resulting from combinations of normalization and dimension reductions. \textbf{B:} Clustering accuracy, mean proportion of the variance in the first 5 components explained by real subpopulations (center), and median ARI of the resulting clustering (right).}
    \label{fig:figure7}
\end{figure}

@@

A common step following dimension reduction is the selection of an appropriate number of dimensions to use for downstream analysis. Since Euclidean distance decreases as the number of non-discriminating dimensions increases, there is usually a trade-off between selecting enough dimensions to keep most information and excluding smaller dimensions that may represent technical noise or other unwanted sources of variation. Overall, increasing the number of dimensions robustly led to a decrease in the number of clusters (Supplementary Figures 10 and 15). This tended to be translated to the accuracy of the clustering (Supplementary Figure 16), although in both cases (number of clusters and ARI) \textit{Seurat}'s resolution parameter had a much stronger impact.

@@

Since ARI is dominated by differences in the number of clusters (Supplementary Figure 17), and no single metric is perfect, we diversified them (Figure \ref{fig:figure9}). Mutual information (MI) has the virtue of not decreasing when a true subpopulation is split into two clusters, which is arguably less problematic (and might well reflect unknown biological subgroups), but as a consequence it can be biased towards methods producing higher resolution clustering. Similarly, precision per true subpopulation is considerably more robust to differences in the number of clusters. We also tracked the mean F1 score and the ARI at the true number of clusters.

The MI score and minimum precision, which are largely independent of the estimated number of clusters, were overall higher for the \textit{walktrap} method (Figure \ref{fig:figure9}), while the mean F1 score favored both \textitt{scran} methods (\textit{walktrap} and \textit{fast greedy}) over \textit{Seurat}. Finally, the ARI score at the true number of clusters, when available, showed similar performances. However, while \textit{Seurat}'s resolution parameter had a large impact on the number of clusters identified, and therefore could be used to find a parameter combination producing the right number of clusters, the number of clusters found by \textit{scran}-based clustering was considerably less influenced by the available parameters (number of nearest neighbors or steps in the random walk - see Supplementary Figure 18). The occasional lack of \textit{scran}-based clustering with the right number of clusters, along with a higher mutual information, suggest that \textit{scran} sometimes simply divides a real subpopulation into two clusters (possibly tracking some unknown biological differences) rather than committing misclassification errors. Overall, the \textit{walktrap} method appeared superior to the \textit{fast greedy} algorithm and was generally less prone to misclassification than \textit{Seurat} clustering, although the latter offered more control over the resolution. Of note, very difficult subpopulations, both from the Zhengmix8eq and simMix1 datasets, remained very inaccurately classified by all methods and in regards to all metrics.

@@

The basic pipeline presented here can be extended with additional analysis steps while keeping the same evaluation metrics. To demonstrate this, we evaluated various imputation or denoising techniques based on their impact on classification. Since preliminary analysis showed that all methods performed equally well or better on normalized data, we applied them after filtering and normalization, but before scaling and reduction. Although some of the methods (e.g. \textit{DRImpute\_process} and \textit{alra\_norm}) did improve the separability of some more elusive subpopulations, no method had a systematically positive impact on the average silhouette width across all subpopulations (Figure \ref{fig:figure10}A). When restricting ourselves to clustering analyses that yielded the `right' number of clusters, all tested methods improved classification  compared to a scenario with no imputation step (`none' label, Figure \ref{fig:figure10}B). However, the situation was not so straightforward with alternative metrics, where some methods (e.g. \textit{enhance}) consistently underperformed. 10X datasets, which are typically characterized by a lower per-cell coverage and feature detection rate, benefited more from imputation, and in this context, \textit{DrImpute} and \textit{dca} tended to show the best performance. On the contrary, imputing on normalized counts originating from Smart-seq technology was instead rather deleterious to the clustering accuracy. The good performance of \textit{DrImpute} is in line with a previous study focused on the preservation of data structure in trajectory inference from scRNAseq \citep{ZhangImput2018} (\textit{alra} was however not included in this previous benchmark).

@@

\begin{enumerate}
   \item Filtering
   \begin{itemize}
     \item Doublet detection and removal is advised, and can be performed at little computing cost with software such as \textit{scDblFinder} or \textit{scds}.
     \item Distribution-based cell filtering fails to capture doublets, and should use relatively lenient cutoffs (e.g. 5 MADs, or 3 MADs in at least 2 distributions) to exclude poor-quality cells while avoiding bias against some subpopulations.
     \item Features filtering based on feature type did not appear beneficial.
   \end{itemize}
   \item Normalization and scaling
   \begin{itemize}
     \item Most normalization methods tested yielded a fair performance, especially when combined with scaling, which tended to have a positive impact on clustering.
     \item \textit{sctransform} offered the best overall performance in terms of the separability of the subpopulations, as well as removing the effect of library size and detection rate.
     \item The common practice of regressing out cell covariates, such as the detection rate or proportion of mitochondrial reads nearly always had a negative impact, leading to increased correlation with covariates and decreased clustering accuracy. We therefore advise against this practice.
   \end{itemize}
   \item Feature selection
   \begin{itemize}
     \item Deviance \citep{townesGlmpca2019} offered the best ranking of genes for feature selection.
     \item Increasing the number of features included tended to lead to better classifications, plateauing from 4000 features.
   \end{itemize}
   \item Denoising/imputation
   \begin{itemize}
       \item Denoising appeared beneficial to the identification of subpopulations in 10x datasets, but not in Smart-seq datasets.
       \item We found especially \textit{alra} (with prior normalization), \textit{DrImpute} (with prior processing) and \textit{dca} to offer the best performances.
   \end{itemize}
   \item PCA
   \begin{itemize}
     \item Similarly to previous reports \citep{SunDimRed2019}, we recommend the irlba-based PCA using weighting of the components, as implemented in \textit{Seurat}. 
     \item Instead of the common elbow or jackstraw methods for deciding on the components to include, we recommend the \textit{global maximum likelihood based on translated Poisson Mixture Model} method (e.g. implemented in \texttt{intrinsicDimension::maxLikGlobalDimEst()}).
   \end{itemize}
   \item Clustering
   \begin{itemize}
    \item We found \textit{scran}-based \textit{walktrap} clustering to show good performances.
    \item In cases where prior knowledge can guide the choice of a resolution, \textit{Seurat} can be useful in affording manual control of it while, in the absence of such knowledge, \textit{scran}-based \textit{walktrap} clustering provided reasonable estimates.
   \end{itemize}
\end{enumerate}

@@

In this study, we evaluated tools commonly used for the processing of scRNAseq with a focus on droplet-based datasets, namely from the 10x technology. Although this platform has been used in almost half of the scRNAseq studies in 2019 \citep{SvenssonDB2019}, other popular technologies such as Drop-seq, InDrops or Smart-seq2/3 were not represented in the present benchmarking. Differences between such protocols can have a very large impact on clustering \citep{MereuCellAtlas2019}. Although most top-ranking methods in our comparison performed well on both Smart-seq and 10x datasets that we tested, future benchmarking efforts should strive to include less represented technologies. In addition, we did not compare any of the alignment and/or quantification methods used to obtain the count matrix, which was for instance discussed in \citep{viethSystematic2019}. Some steps, such as the implementation of the PCA, were also not explored in detail here as they have already been the object of recent and thorough study elsewhere \citep{SunDimRed2019, TsuyuzakiPCA2020}. We also considered only methods relying on Euclidean distance, while correlation was recently reported to be superior\citep{kim_impact_2019} and would require further investigation.

Here, we chose to concentrate on what could be learned from datasets with known cell labels (as opposed to labels inferred from the data, as in \citep{MereuCellAtlas2019}). In contrast to Tian et al. \citep{tianMixology2018}, who used on RNA mixtures of known proportions, we chose to rely chiefly on real cells and their representative form of variability. Given the limited availability of such datasets however, several aspects of single-cell analysis could not be compared, such as batch effect correction or multi-dataset integration. For these aspects of scRNAseq processing that are critical in some experimental designs, we refer the reader to previous evaluations \citealp{TranBatch2020, StuartIntegration2019}. In addition, a focus on the identification of the subpopulations might fail to reveal methods that are instead best performing for tasks other than clustering, such as differential expression analysis or trajectory inference. Several informative benchmarks have already been performed on some of these topics \citealp{CrowellMuscat2019, DalMolinDE2017, JaakkolaDE2017, SaelensTraject2019, SonesonDE2018, WangDE2019}. Yet, such evaluations could benefit from considering methods not in isolation, but as parts of a connected workflow, as we have done here. We believe that the \textit{pipeComp} framework is modular and flexible enough to integrate new steps in the pipeline, as shown by an example with imputation/denoising methods.

@@

\textit{pipeComp} is a flexible R framework for evaluating methods and parameter combinations in a complex pipeline, computing on-the-fly multilevel evaluation metrics. Applying this framework to scRNAseq clustering enabled us to make concrete recommendations on the steps of filtering, normalization, feature selection, denoising, dimensionality reduction and clustering. We demonstrate how a diversity of multilevel metrics can be more robust, more sensitive, and more nuanced than simply evaluating the final clustering performance. In addition, we provide a new and efficient Bioconductor package for doublet detection, \texttt{scDblFinder}. We hope that the \textit{pipeComp} framework can be applied to extend the current field of benchmarking, as well as to apply it to other range of methods.

@@

\subsubsection*{Denoising/imputation}
Most imputation/denoising methods were run with the default parameters with the following exceptions; \textit{DrImpute} documentation advises to process the data prior to imputation by removing lowly expressed genes and cells expressing less than 2 genes. As it is not clear if this step is a hard requirement for the method to perform well, \textit{DrImpute} was run with and without prior processing (\textit{DrImpute\_process} and \textit{DrImpute\_noprocess} labels, respectively). The \textit{dca} method accepts only integers, while two of the datasets had non-integer quantification of expected counts. For these datasets, we rounded up the counts prior to imputation. \textit{alra} is designed for normalized data but as we are evaluating normalization downstream to imputation, we used the method on both non-normalized (\textit{alra} label) and normalized counts (\textit{alra\_norm} label). \textit{scImpute} requires an estimation of the expected number of clusters with the input data. As the estimation of the true number of cluster may not be known by the user, we evaluated the tool using the true number of clusters (\textit{scImpute} label) and using an over/under-estimation of this number (\textit{scImpute\_plus5} and \textit{scImpute\_min5} labels, respectively). \textit{ENHANCE} uses a k-nearest neighbor aggregation method and automatically estimates the number of neighbors to merge prior to the imputation. With the smallest datasets, this parameters was estimated to be 1, which lead to an early stop of the function. In such cases, we manually set this parameter to 2 for the method to work.
